scRNAseq benchmark in want of a title

### 2019-06-03 bmc_article.tex
mark edits jun 3

Although we use some datasets based on other protocols, our focus is especially on 10X datasets and on methods that do not include exogenous control RNA, i.e. spike-ins (see Table \ref{tab:table1} and Supplementary Figure 1 for a description of the datasets). As Seurat \cite{satijaSeurat2015} was previously shown to outperform other clustering approaches \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting point to perform an integrated investigation of 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimensional reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameter and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

We define a pipeline, in R, as a list of functions executed consecutively on the output of the previous one (Figure 0). Each step can include various parameters, all of which must be given alternative values. The \textit{runPipeline} command will then proceed in a step-wise fashion to computer only the parts of the pipelines that have not already been computed. In addition, each step can output evaluation metrics, which together provide a multi-layered evaluation, where variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of cluster \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and as we will see they might not always capture improvements in earlier steps of the pipeline.

@@

In addition to these methods, we reasoned that an approach such as DoubletFinder could be considerably sped up by being applied directly on counts and by using a pre-clustering to create neotypic doublets more efficiently. We therefore developed a simple and fast package implementing this for doublet detection, with the added perk of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).

@@

\subsubsection*{Excluding more cells is not necessarily good}

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for cells that have a high content of mitochondrial reads, often as a result of cell degradation [REF]. A common practice is to exclude cells that differ considerably from most other cells on the basis of some such properties, for instance through the `isOutlier` function of `scater` which measures, for a given control property, the number of median absolute deviation of each cell from the mean of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, but are not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3). In addition, in our experience 10X datasets also exhibit a very tight correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to eliminate are cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation, and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7). We could not identify any property or simple combination thereof which would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with a very high read counts have a higher chance of being misclassified.

@@

We next investigated the contribution of various criteria individually (Supplementary Figure 9). Interestingly, regression analysis suggests that increasing the number of distributions on which a cell must deviate in order to be excluded tended to be positively associated with ARI, while increasing the number of median absolute deviations required to be an outlier as associated with a decrease in ARI. We therefore suggest excluding cells that are outliers on more than 1 distributions. The impact of filtering on the basis of total counts or feature counts was not always robust, nor always positive, and we could detect no significant impact of filtering on the basis of the proportion of mitochondrial reads - however it must be said that the benchmark datasets count very few cells with a very high proportion of mitochondrial reads (Supplementary Figure 1). We therefore suggest to keep this criterion, but in a relatively lenient fashion, and not solely based on the median absolute deviation of the distribution. Filtering solely based on deviations from the distribution might be inadequate when the distribution is tight; in our experience, the proportion of mitochondrial counts in otherwise `healthy` cells can vary up to 8-12\% depending on the tissue. We therefore recommend using a combination of fixed and deviation based threshold, e.g. x $>$ 0.08 and x $>$ 2.5*MADs.

@@

On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further. We found that filtering had a much milder impact than doublet removal (Figure \ref{fig:figure3}), and that more stringent filtering tended to result in higher silhouette width, in a higher proportion of the variance in the first components explained by subpopulations (Supplementary Figure 10), and often in a higher adjusted Rand index (ARI) of the clustering at the right number of clusters (Figure \ref{fig:figure3}). However, part of the gain is to be ascribed to the removal of cells from subpopulations (especially in the Zheng datasets) which are difficult to distinguish, and for which more stringent clustering did not improve silhouette width (Supplementary Figure 10A). For these reasons, we therefore recommend using a more lenient filter (e.g. lenient or default filter described here) or a cluster-based filter to avoids most of the subpopulation bias while excluding strong outliers.

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in Seurat, we tested scran's pooling-based normalization \citep{lunPooling2016}, sctransform's variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}. 
In addition to log-normalization, the standard Seurat clustering pipeline scales features so that the PCA is not dominated by a small subset of the features, and we therefore included versions of the different normalization, with or without a subsequent scaling (except sctransform's variance-stabilizing transformation, given that, conceptually, it involves something analogous to scaling). 
Seurat's scaling function includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features. Finally, it has been proposed that the use of stable genes, and in particular cytosolic ribosomal genes, can be used to normalized scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

>>>>

Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets and on methods that do not include exogenous control RNA, i.e. spike-ins (see Table \ref{tab:table1} and Supplementary Figure 1 for a description of the datasets). Since graph-based clustering  \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameter and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

We define a pipeline, in R, as a list of functions executed consecutively on the output of the previous one (Figure 0). Each step can include various parameters, all of which must be given alternative values. The \textit{runPipeline} command will then proceed in a step-wise fashion to compute only the parts of the pipelines that have not already been computed. In addition, each step can output evaluation metrics, which together provide a multi-layered evaluation, where variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of cluster \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and as shown below, might not always capture improvements in earlier steps of the pipeline.

@@

In addition to these methods, we reasoned that an approach such as DoubletFinder could be considerably sped up by being applied directly on counts and by using a pre-clustering to create neotypic doublets more efficiently. We therefore developed a simple and fast package implementing this for doublet detection, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).

@@

\subsubsection*{Excluding more cells is not necessarily better}

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for cells that have a high content of mitochondrial reads, often as a result of cell degradation [REF]. A common practice is to exclude cells that differ considerably from most other cells on the basis of some such properties, for instance through the `isOutlier` function of `scater` which measures, for a given control property, the number of median absolute deviation of each cell from the mean of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, but not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3). In addition, in our experience 10X datasets also exhibit a very tight correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to eliminate are cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation, and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with a very high read counts have a higher chance of being misclassified.

@@

We next investigated the contribution of various criteria individually (Supplementary Figure 9). Interestingly, regression analysis suggests that increasing the number of distributions on which a cell must deviate in order to be excluded tended to be positively associated with ARI, while increasing the number of median absolute deviations required to be an outlier as associated with a decrease in ARI. We therefore suggest excluding cells that are outliers in more than 1 distribution. The impact of filtering on the basis of total counts or feature counts was not always robust, nor always positive, and we could detect no significant impact of filtering on the basis of the proportion of mitochondrial reads - however it must be said that the benchmark datasets count very few cells with a very high proportion of mitochondrial reads (Supplementary Figure 1). We therefore suggest to keep this criterion, but in a relatively lenient fashion, and not solely based on the median absolute deviation of the distribution. Filtering solely based on deviations from the distribution might be inadequate when the distribution is tight; in our experience, the proportion of mitochondrial counts in otherwise `healthy` cells can vary up to 8-12\% depending on the tissue. We therefore recommend using a combination of fixed and deviation based threshold, e.g. x $>$ 0.08 and x $>$ 2.5*MADs.

@@

On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further. We found that filtering had a much milder impact than doublet removal (Figure \ref{fig:figure3}), and that more stringent filtering tended to result in higher silhouette width, in a higher proportion of the variance in the first components explained by subpopulations (Supplementary Figure 10), and often in a higher adjusted Rand index (ARI) of the clustering at the right number of clusters (Figure \ref{fig:figure3}). However, part of the gain is to be ascribed to the removal of cells from subpopulations (especially in the Zheng datasets) that are difficult to distinguish, and for which more stringent clustering did not improve silhouette width (Supplementary Figure 10A). For these reasons, we therefore recommend using a more lenient filter (e.g. lenient or default filter described here) or a cluster-based filter to avoid most of the subpopulation bias while excluding strong outliers.

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in Seurat, we tested scran's pooling-based normalization \citep{lunPooling2016}, sctransform's variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}. 
In addition to log-normalization, the standard Seurat clustering pipeline scales features so that the PCA is not dominated by a small subset of the features, and we therefore included versions of the different normalization, with or without a subsequent scaling (except sctransform's variance-stabilizing transformation, given that, conceptually, it involves something analogous to scaling). 
Seurat's scaling function includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features. Finally, it has been proposed that the use of stable genes, and in particular cytosolic ribosomal genes, can be used to normalize scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
